--- 
title: "ml4ec - Machine Learning for Eddy Covariance data"
author: "Benjamin Stocker"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is a tutorial for a workshop (1 day), introducing machine learning using R."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.align  = "center",
  fig.width = 7.25,
  fig.height = 6
  )
system("mkdir data")
```

# Set up

## Apps

For this workshop, you need [R](https://www.r-project.org/) and [RStudio](https://www.rstudio.com/products/rstudio/download/). Follow the links for downloading and installing these apps.

## Libraries

Install missing packages for this tutorial.
```{r}
list_pkgs <- c("caret", "recipes", "rsample", "tidyverse", "conflicted", "modelr", "forcats", "yardstick", "visdat", "skimr", "ranger", "knitr")
new_pkgs <- list_pkgs[!(list_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs) > 0) install.packages(new_pkgs)
```

This book was compiled with the *bookdown* library and source files (RMarkdown), available on [Github](https://github.com/stineb/ml4ec_workshop). Navigate there also for working on the exercises (Chapter \@ref(exercises)) and using the solutions (Chapter \@ref(solutions)).

This book is published under a GNU-GPL v3 license. Please cite this as:

*xxxx*

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
